### 2024-09-20 MetadataS1/MetadataS1.tex
prepare manuscript for Ecology journal

%% Language %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[top=2.5cm, bottom=2.5cm, left=2.5cm, right=2.5cm]{geometry}
\usepackage{scrextend}
\addtokomafont{labelinglabel}{\sffamily}
\usepackage{lmodern} %Type1-font for non-english texts and characters
\usepackage{parskip}

%% Packages for Graphics, Figures & Tables %%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{graphicx} %%For loading graphic files
\usepackage{subcaption}
\usepackage{float}
\usepackage{booktabs} %% for top/midrule
\usepackage{textcomp}
\usepackage[table,xcdraw]{xcolor}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{makecell}
\usepackage{pdflscape}
%bibliography
\usepackage[round]{natbib}
\bibliographystyle{abbrvnat}
\usepackage{hyperref}
\hypersetup{
    colorlinks=false,
    pdfborder={0 0 0} % Removes the border
}
\usepackage{geometry}

@@

@@

\begingroup  
  \centering
\textbf{Long-term abundance time-series of the High Arctic terrestrial vertebrate community of Bylot Island, Nunavut}\\[1.5em]
\endgroup

@@

\textbf{Author Affiliations}:\\
\textsuperscript{1} Chaire de Recherche du Canada en Biodiversité Nordique, Centre d’Études Nordiques, Centre de la science de la biodiversité du Québec, Département de biologie, chimie et géographie, Université du Québec à Rimouski, Rimouski, QC, Canada\\
\textsuperscript{2} Chaire de Recherche du Canada en Écologie Intégrative, Centre d’Études Nordiques, Centre de la science de la biodiversité du Québec, Département de Biologie, Université de Sherbrooke, Sherbrooke, QC, Canada\\
\textsuperscript{3} Chaire de Recherche Sentinelle Nord sur l’impact des migrations animales sur les écosystèmes nordiques, Centre d’Études Nordiques, Centre de la science de la biodiversité du Québec, Département de Biologie, Université Laval, Québec, QC, Canada\\
\textsuperscript{4} Centre d’études nordiques, Département de Biologie, Université Laval, Québec, QC, Canada\\
\textsuperscript{5} Centre d’études nordiques, Centre de connaissance et d’exploration de l’Arctique, Musée canadien de la nature, Ottawa, ON, Canada\\
\textsuperscript{6} Centre d’Études Biologiques de Chizé (CEBC-CNRS), Université de La Rochelle, France 
\\[1.5em]

\textbf{Corresponding Authors}:\\
\textit{\href{mailto:louis.moisan.bio@gmail.com}{louis.moisan.bio@gmail.com}}\\
\textit{\href{mailto:joel_bety@uqar.ca}{joel{\textunderscore}bety@uqar.ca}}
\\[1.5em]

\section*{Abstract}
Arctic ecosystems present unique opportunities for community-wide monitoring, in part due to their relatively low species richness. However, conducting research in these remote environments poses significant logistical challenges, resulting in long-term monitoring being exceedingly rare. Here, we focus on the long-term, intensive ecological monitoring efforts conducted on the south plain of Bylot Island (almost 400 km\textsuperscript{2}, Nunavut, Canada), which has generated a remarkable dataset spanning up to 30 years, a rarity in tundra ecosystems. Our goal is to synthesize this dataset and upscale vertebrate abundance data at the landscape level, a prerequisite to conduct community-level analyses. We have standardized data obtained with different field methods to provide readily usable long-term time series of abundance for 35 vertebrate species (30 birds and 5 mammals) present in the study system. Monitoring data includes intensive capture-mark-recapture density estimates of lemmings on trapping grids, systematic or opportunistic nest monitoring conducted across the entire study area or within specific plots for all bird species, transects of vertebrate counts distributed throughout the study area, daily incidental observations of vertebrates and satellite tracking of fox movements. Annual abundance of species was estimated at the landscape level, accounting for spatial variations. Furthermore, we provide body masses for each species, derived from empirical onsite measurements for 18 species and from the literature for the remaining species. Body mass is essential to convert species abundance into biomass for studies of trophic fluxes and ecosystem processes. Our dataset provides a unique opportunity for holistic empirical studies of ecological communities, allowing a deeper understanding of community structure and dynamics. Considering that the study site is a pristine and protected area that has experienced minimal anthropogenic impact, it can also provide an ideal baseline for investigating the impacts of global changes on high-latitude terrestrial ecosystems.

\section*{Introduction}
The composition of ecological communities, defined as the abundance of each species within a given community, is fundamental for understanding patterns and processes in community ecology. Variations in community composition can help to detect spatial patterns linked to environmental variations \citep{kemp1990}, assess temporal trends of different groups of species after disturbances \citep{philippi1998, magurran2007}, and understand food web structures \citep{cohen2003}. Additionally, community composition is essential for modeling the dynamics of ecological communities. Dynamic community modelling allows addressing important issues and questions in ecology, such as: determining the relative strength of top-down versus bottom-up forces in communities \citep{krebs2003,legagneux2014}, assessing the ecological resilience of communities under climate change \citep{griffith2019} and evaluating the cascading effects of invasive species in food web \citep{david2017, goto2020}. Dynamic community modelling can also be applied to address practical challenges, including fishery management (Plagányi, 2007) and the planning of protected areas \citep{okey2004, dahood2020}.

@@

\section{Study area}
The study area (389 km\textsuperscript{2}) is located on the south plain of Bylot Island, Nunavut, Canada. It comprises a combination of mesic tundra mainly on hills (64 \%), upland plateaus of sedimentary rock with drier/rockier habitat at higher elevation (20 \%), low-lying wetlands interspersed with ponds (10 \%) and larger bodies of water such as lakes and rivers (6 \%). Wetlands were delineated by photo-interpretation of high-resolution satellite images (30 cm; Ouellet, unpublished data), whereas lakes were delineated with aerial photos and rivers with google satellite images, resulting in a coarser delineation. The mean temperature in July is 6\textdegree C, and the area typically remains free of snow from mid-June to late September \citep{gauthier2013}. An important biological characteristic of the area is the presence of a large snow goose (scientific names of most vertebrate species can be found in \textbf{Table \ref{table:species_name_strategy}}) colony of around 25 000 breeding pairs \citep{reed2002} spanning approximately 70 km\textsuperscript{2}. The vertebrate community within the study area comprises 30 bird species, with 29 of them being migratory or partially migratory, along with 5 mammal species (\textbf{Table \ref{table:species_name_strategy}}; \citet{moisan2023, gauthier2024a}). The study area experiences significant temporal fluctuations in the population of small mammals (lemmings), which in turn impact the occurrence and abundance of their avian and mammalian predators such as snowy owls, rough-legged hawks, long-tailed jaegers, ermines and Arctic foxes \citep{legagneux2012}. We exclude occasional visitors, namely: i) species lacking confirmed breeding occurrences on the study site, ii) species observed solely within a single year, and iii) species primarily breeding and foraging in nearby marine or coastal habitats \citep{moisan2023}. The case of the red fox (Vulpes vulpes) was ambiguous. While the presence of a breeding pair has been confirmed in the study area \citep{lai2022}, the extent of population establishment remains unclear and sightings are rare. Therefore, we decided to exclude this species.

The study area is divided into 9 zones based on the sampling method and the level of field effort applied in each zone (\textbf{Figure \ref{figure:zones}}). Long-term monitoring of the community began in the 1990s in the Qarlikturvik valley \citep{gauthier2013, gauthier2024a}, which represents the zone of the study area with the highest annual sampling effort. Within the Qarlikturvik valley, the sampling is concentrated on the southern side of the glacial river, where the main research infrastructure is located. Another zone with extensive sampling efforts is Camp 2, located at the core of the snow goose colony, where the primary focus is to monitor snow goose nests. However, nests of many other avian species are also monitored within and around the colony in this zone. Camp 3, Pointe Dufour, Goose Point, and Malaview are zones where intensive sampling efforts are conducted annually, albeit for a relatively brief period (approximately one week) during the breeding season of most species \citep{gauthier2024a}. The upland zones in the study area (defined as areas approximately 300 meters above sea level or more) are the Black Plateau, Southern Plateau, and Camp 3 Plateau. These zones are primarily visited to assess raptor nesting activity \citep{beardsell2016}. The zone between the Qarlikturvik valley and Camp 3 received very little sampling effort and is therefore excluded from the study area.

\begin{figure}
\centering
  \includegraphics[width=0.8\textwidth, angle=0]{figures/zones_study_area.pdf}
   \vspace{-35pt} % Reduce space between figure and caption
   \caption{Map of the different zones (colored polygons) of the 389 km\textsuperscript{2} study area located on the south plain of Bylot Island, Nunavut Canada. The perimeter of the snow goose colony is delineated by white dashes; we highlighted the perimeter in 2017 since it represents the average colony area.}
  \label{figure:zones}
\end{figure}
\newpage

\input{tables/species_name_strategy.tex}
%label-> table:species_name_strategy
\newpage

\section{Data collection}
\subsection{Avian nest monitoring}
Avian nest monitoring was conducted annually except in 2020 and 2021 due to logistical constraints imposed by the COVID-19 pandemic. Systematic nest monitoring refers here to a systematic sampling approach aimed at documenting all nests within a specified area. Monitoring is considered opportunistic when there is a chance that some nests might not have been detected within a specific area. Nest densities derived from nest sampling could be underestimated due to early nest failure (i.e., failure that happened before our sampling period).

\input{tables/species_year_monitoring.tex}
%label-> table:species_year_monitoring
\newpage

\subsubsection{Systematic nest monitoring}
\subsubsection*{Pacific loon, red-throated loon, cackling goose, tundra swan and glaucous gull}
Since 2004, systematic searches of wetland areas have been conducted on the southern side of the glacial river in the Qarlikturvik Valley, and since 2017, in other zones of the study area. This sampling aimed to find all nests of the cackling goose and the glaucous gull. Nest locations of other large wetland-nesting species, including the tundra swan, the red-throated loon and the Pacific loon, were also noted, as these species nest in the same habitat \citep{duchesne2021,gauthier2024a}. Each year, all known or potential nesting sites were revisited. Observers detected nests by walking and scanning around ponds and lakeshores to identify any active nesting sites. These large species can be seen from a relatively long distance sitting on the nest or when flushing from the nest. Most of them (geese, swans and gulls) can also reveal their presence with alarm calls or nest defense displays. We are confident that nest detection probability was high for these species given the open landscape.

\subsubsection*{Snow goose}
Snow geese nest in a large colony in the study area (\textbf{Figure \ref{figure:zones}}), but also in small aggregations distributed on the island, especially in years when snowy owls are nesting \citep{lepage1996,reed2002}. Since 1994, goose nests were systematically monitored on a 0.24 km\textsuperscript{2} wetland at the center of the colony. Since 1999, nests were also systematically monitored on a variable number of plots, measuring 0.01 km\textsuperscript{2} in wetland habitat and 0.04 km\textsuperscript{2} in mesic habitat, randomly distributed throughout the goose colony \citep{gauthier2020goose}. The total area covered by the randomly distributed plots averaged 0.79 \textpm{0.37} km\textsuperscript{2} per year. From 2010 onwards, except in 2020 and 2021, we traced the approximate boundary of the goose colony using a GPS receiver aboard a helicopter flying along the colony border \citep{duchesne2021}.

\subsubsection*{Rough-legged hawk, peregrine falcon and common raven}
Peregrine falcons, rough-legged hawks and common ravens nest on cliffs, near ravines, and on large rocky outcrops and tend to reuse the same nesting sites from one year to the next \citep{beardsell2016}. Systematic monitoring of every known or potential nesting site has been carried out in the Qarlikturvik valley, Black plateau and Southern plateau since 2007 and throughout the study area since 2013 \citep{beardsell2016, gauthier2020avianpred}. Observers walked along ridges and scanned surrounding areas from vantage points to detect nesting birds. These large species can be seen from a relatively long distance sitting on the nest or when flushing from the nest. They can also reveal their presence with alarm calls or nest defense displays. We are confident that nest detection probability was high for these species. Each year the observers use slightly different paths to sample the areas, but locate the nests in the same positions, which supports a high probability of detection for these species. Most nesting sites were located in the upland zones of the study area, which include the Black Plateau, Southern Plateau and Camp 3 Plateau.

\subsubsection*{Snowy owl}
Snowy owls predominantly nest in habitats similar to other raptors, favoring ridges in mountainous or hilly regions, although they can occasionally be found nesting on mounds in lowland areas \citep{seyer2020}. Since 1996, searches for snowy owl nests have been conducted concurrently with searches for other raptor nests in the Black and Southern plateaus, as well as during searches for jaeger nests on the southern side of the glacial river in the Qarlikturvik Valley. Additionally, since 2012, nests have been recorded across the entire study area by scanning the landscape from hills and ridges during the nesting period \citep{duchesne2021}. Given that snowy owls nest on elevated mounds, exhibit contrasting colors with the landscape, emit alarm calls, and display defensive behaviors, active nesting sites have a high probability of detection.

\subsubsection*{Long-tailed jaeger and parasitic jaeger}
Since 2004, observers have walked parallel transects spaced 400 meters apart, covering the entire southern side of the glacial river in the Qarlikturvik Valley (33 km\textsuperscript{2}; \textbf{Figure \ref{figure:qarlikturvik_valley}}), during the nesting period. The aim of those transects was to record nests of long-tailed jaegers, parasitic jaegers, and sandhill cranes. Observers listened for alarm calls to detect territorial birds, and then located nests by observing the birds returning to their nests from elevated vantage points. We consider the sampling to be systematic for long-tailed and parasitic jaeger, since those species tend to leave their nest relatively far from the observer to perform mobbing behavior, and thus increasing their detection probability. We do not consider the sampling to be systematic for sandhill cranes as they only display defensive behaviors near their nests at relatively short distances (see opportunistic nest monitoring below).

\begin{figure}
\centering
  \includegraphics[width=0.8\textwidth, angle=0]{figures/qarlikturvik_valley.pdf}
  \vspace{-70pt} % Reduce space between figure and caption
  \caption{Intensive nests searching plots (8 km\textsuperscript{2} and 2 km\textsuperscript{2}) located in the Qarlikturvik valley.}
 \label{figure:qarlikturvik_valley}
\end{figure}

\subsubsection*{Common-ringed plover}
Between 2015 and 2019, observers conducted surveys of the primary nesting areas of the common-ringed plover. The survey involved walking in stony and sandy shores and gravel bars with scarce vegetation along rivers. Nests were found by detecting individuals exhibiting reproductive behaviors, such as incubation, alarm calls, or distraction displays. The sampling effort was particularly intensive between 2015 and 2017. Small areas along the coast or on the banks of smaller rivers that could potentially serve as nesting sites may have been overlooked.

\subsubsection*{Lapland longspur and Baird's sandpiper}
Since 2005, nests of passerines and sandpipers have been extensively monitored across an 8 km\textsuperscript{2} (4x2 km) area in the Qarlikturvik valley. We considered the sampling to be most systematic within a core 2 km\textsuperscript{2} (2x1 km) plot in this area (\textbf{Figure \ref{figure:qarlikturvik_valley}}). We excluded relatively large water bodies (0.26 km\textsuperscript{2}) to calculate nest density in the plot due to the presence of a large lake, which leaves an area of 1.74 km\textsuperscript{2} available for nesting. An observer conducted systematic searches of this plot during the entire breeding season to locate and monitor as many passerine and shorebird nests as possible. Assuming the observer can detect all nests within a 5 or 10 meter radius, analysis of daily GPS tracks shows that the observer covered a minimum area of 0.72 \textpm{0.12} (5 m) or 1.09 \textpm{0.17} km\textsuperscript{2} (10 m) of the core area annually (n= 3 years). Additionally, several other observers conducting related field work in the same zone reported all passerine and shorebird nests found opportunistically.

>>>>

%% Language %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[top=2.5cm, bottom=2.5cm, left=2.5cm, right=2.5cm]{geometry}
\usepackage{scrextend}
\addtokomafont{labelinglabel}{\sffamily}
\usepackage{lmodern} %Type1-font for non-english texts and characters
\usepackage{parskip}
\usepackage{enumitem}

%% Packages for Graphics, Figures & Tables %%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{graphicx} %%For loading graphic files
\usepackage{subcaption}
\usepackage{float}
\usepackage{booktabs} %% for top/midrule
\usepackage{textcomp}
\usepackage[table,xcdraw, dvipsnames]{xcolor}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{makecell}
\usepackage{pdflscape}

%bibliography
\usepackage[round]{natbib}
\bibliographystyle{abbrvnat}
\usepackage{hyperref}
\hypersetup{
    colorlinks=false,
    pdfborder={0 0 0} % Removes the border
}
\usepackage{geometry}

@@

\usepackage{titlesec}
\titlespacing{\section}{0pt}{*4}{*1.5} 
\titlespacing{\subsection}{20pt}{*4}{*1.5}
\titlespacing{\subsubsection}{40pt}{*4}{*1.5} 
\titlespacing{\paragraph}{60pt}{*4}{*1.5}
\titlespacing{\subparagraph}{80pt}{*4}{*1.5}

@@

\begingroup  
  \centering
\textbf{Title: Long-term abundance time-series of the High Arctic terrestrial vertebrate community of Bylot Island, Nunavut}\\[1.5em]

@@

\textbf{Author Affiliations}:\\
\textsuperscript{1} Chaire de Recherche du Canada en Biodiversité Nordique, Centre d’Études Nordiques, Centre de la science de la biodiversité du Québec, Département de biologie, chimie et géographie, Université du Québec à Rimouski, Rimouski, QC, Canada\\
\textsuperscript{2} Chaire de Recherche du Canada en Écologie Intégrative, Centre d’Études Nordiques, Centre de la science de la biodiversité du Québec, Département de Biologie, Université de Sherbrooke, Sherbrooke, QC, Canada\\
\textsuperscript{3} Chaire de Recherche Sentinelle Nord sur l’impact des migrations animales sur les écosystèmes nordiques, Centre d’Études Nordiques, Centre de la science de la biodiversité du Québec, Département de Biologie, Université Laval, Québec, QC, Canada\\
\textsuperscript{4} Centre d’études nordiques, Département de Biologie, Université Laval, Québec, QC, Canada\\
\textsuperscript{5} Centre d’études nordiques, Centre for Arctic Knowledge and Exploration, Canadian Museum of Nature, Ottawa, ON, Canada\\
\textsuperscript{6} Centre d’Études Biologiques de Chizé (CEBC-CNRS), Université de La Rochelle, France 
\\[1.5em]

\textbf{Corresponding Author}:\\
\textit{\href{mailto:louis.moisan.bio@gmail.com}{louis.moisan.bio@gmail.com}}\\
\textit{\href{mailto:joel_bety@uqar.ca}{joel{\textunderscore}bety@uqar.ca}}
\\[1.5em]

\textbf{Open Research statement}:

%%%%%%% !!!! ADD DRYAD adress !!!! %%%%%%%%
Data are publicly available at \url{https://datadryad.org/}.\\[1.5em]
Raw data and codes used to extract the presented data set are publicly available at \url{https://github.com/chaireBioNorth/BYLOT_species_abundance_dataset}.
\endgroup
\newpage

\subsection*{Introduction}
 The composition of ecological communities, defined as the abundance of each species within a given community, is fundamental for understanding patterns and processes in community ecology. Variations in community composition can help to detect spatial patterns linked to environmental variations \citep{kemp1990}, assess temporal trends of different groups of species after disturbances \citep{philippi1998, magurran2007}, and understand food web structures \citep{cohen2003}. Additionally, community composition is essential for modeling the dynamics of ecological communities. Dynamic community modelling allows addressing important issues and questions in ecology, such as: determining the relative strength of top-down versus bottom-up forces in communities \citep{krebs2003,legagneux2014}, assessing the ecological resilience of communities under climate change \citep{griffith2019} and evaluating the cascading effects of invasive species in food web \citep{david2017, goto2020}. Dynamic community modelling can also be applied to address practical challenges, including fishery management (Plagányi, 2007) and the planning of protected areas \citep{okey2004, dahood2020}.

@@

\section*{\textcolor{Blue}{Class I. Data Set Descriptors}}
    \section*{A. Data set identity} Long-term abundance time-series of the High Arctic terrestrial vertebrate community of Bylot Island, Nunavut
    \section*{B. Data set identification codes}
     \texttt{BYLOT-species\_taxonomy.csv}\\
     \texttt{BYLOT-species\_abundance.csv}\\
     \texttt{BYLOT-mean\_species\_abundance.csv}\\
     \texttt{BYLOT-mean\_species\_body\_mass.csv}\\
     \texttt{BYLOT-interannual\_variation\_nest\_density.csv}\\

    \section*{C. Data set description}
		\subsection*{1. Originators}
		\textbf{Gilles Gauthier}, Centre d’études nordiques, Département de Biologie, Université Laval, Québec, QC, Canada\\
  		\textbf{Joël Bêty},  Chaire de Recherche du Canada en Biodiversité Nordique, Centre d’Études Nordiques, Centre de la science de la biodiversité du Québec, Département de biologie, chimie et géographie, Université du Québec à Rimouski, Rimouski, QC, Canada \\
  		\textbf{Pierre Legagneux}, Chaire de Recherche Sentinelle Nord sur l’impact des migrations animales sur les écosystèmes nordiques, Centre d’Études Nordiques, Centre de la science de la biodiversité du Québec, Centre d’Études Biologiques de Chizé (CEBC-CNRS) Département de Biologie, Université Laval, Québec, QC, Canada

		\subsection*{2. Abstract}
Arctic ecosystems present unique opportunities for community-wide monitoring, in part due to their relatively low species richness. However, conducting research in these remote environments poses significant logistical challenges, resulting in long-term monitoring being exceedingly rare. Here, we focus on the long-term, intensive ecological monitoring efforts conducted on the south plain of Bylot Island (almost 400 km\textsuperscript{2}, Nunavut, Canada), which has generated a remarkable dataset spanning up to 30 years, a rarity in tundra ecosystems. Our goal is to synthesize this dataset and upscale vertebrate abundance data at the landscape level, a prerequisite to conduct community-level analyses. We have standardized data obtained with different field methods to provide readily usable long-term time series of abundance for 35 vertebrate species (30 birds and 5 mammals) present in the study system. Monitoring data includes intensive capture-mark-recapture density estimates of lemmings on trapping grids, systematic or opportunistic nest monitoring conducted across the entire study area or within specific plots for all bird species, transects of vertebrate counts distributed throughout the study area, daily incidental observations of vertebrates and satellite tracking of fox movements. Annual abundance of species was estimated at the landscape level, accounting for spatial variations. Furthermore, we provide body masses for each species, derived from empirical onsite measurements for 18 species and from the literature for the remaining species. Body mass is essential to convert species abundance into biomass for studies of trophic fluxes and ecosystem processes. Our dataset provides a unique opportunity for holistic empirical studies of ecological communities, allowing a deeper understanding of community structure and dynamics. Considering that the study site is a pristine and protected area that has experienced minimal anthropogenic impact, it can also provide an ideal baseline for investigating the impacts of global changes on high-latitude terrestrial ecosystems.

   \section*{D. Key words/phrases}
 Bylot Island, Canadian Arctic, Arctic tundra, 1993-2023, long-term monitoring, biodiversity monitoring, community composition, species abundance, species density, species biomass, species body mass, food web
  \newpage

\section*{\textcolor{Blue}{Class II. Research origin descriptors}}
    \section*{A. Overall project description}
    \subsection*{1. Identity} Understanding the structure and dynamics of Arctic terrestrial vertebrate communities

	\subsection*{2. Originators:}	
  	\textbf{Gilles Gauthier}, Centre d’études nordiques, Département de Biologie, Université Laval, Québec, QC, Canada\\
  		\textbf{Joël Bêty},  Chaire de Recherche du Canada en Biodiversité Nordique, Centre d’Études Nordiques, Centre de la science de la biodiversité du Québec, Département de biologie, chimie et géographie, Université du Québec à Rimouski, Rimouski, QC, Canada \\
  		\textbf{Pierre Legagneux}, Chaire de Recherche Sentinelle Nord sur l’impact des migrations animales sur les écosystèmes nordiques, Centre d’Études Nordiques, Centre de la science de la biodiversité du Québec, Centre d’Études Biologiques de Chizé (CEBC-CNRS) Département de Biologie, Université Laval, Québec, QC, Canada

	\subsection*{3. Period of study} 1989 - continuing

	\subsection*{4. Objectives}
       i) Understand the factors that shape the structure and drive the dynamics of Arctic terrestrial vertebrate communities.\\
       ii) Predict the effects of current global environmental changes on the structure and dynamics of Arctic terrestrial vertebrate communities.

     \subsection*{5. Abstract}
        Arctic terrestrial vertebrate communities present low species richness, making those relatively simple communities ideal for studying ecological patterns and dynamics in terrestrial environments. Although, they present complex networks of interacting species, extreme seasonal changes of environmental conditions and a large portion of migratory species, which make challenging the identification of the key factors that shape their structure and dynamics. In the face of rapid global environmental changes, it is crucial to have a comprehensive understanding of the key processes shaping Arctic terrestrial communities structure and dynamics in order to predict how global changes will impact them in the future. Our research emphasizes long-term biodiversity monitoring, a community-wide perspective and food web modeling to achieve this understanding.

      \subsection*{6. Sources of funding}
Natural Sciences and Engineering Research Council of Canada, Fonds Québécois de Recherche Nature et Technologies, Centre d'études nordiques, Natural Resources Canada (Polar Continental Shelf Program), Network of Centers of Excellence Canada (ArcticNet), Canada First Research Excellence Fund (Sentinel North Program), Polar Knowledge Canada, Environment and Climate Change Canada, Canada Foundation for Innovation, Parks Canada Agency, International Polar Year program of the Government of Canada, Crown–Indigenous Relations and Northern Affairs Canada (Northern Contaminant Program), Duck Unlimited Canada, Kenneth M. Molson Foundation (Kenneth M. Molson Foundation’s donation for wildlife research, conservation, and habitat), Garfield Weston Foundation, First Air——Canadian North, Nunavut Wildlife Management Board, Université Laval, Université du Québec à Rimouski
\newpage

    \section*{B. Specific subproject description}
     	\subsection*{1. Site description}
 			\subsubsection*{a. Site type}
  The study area (389 km\textsuperscript{2}) represents a relatively productive tundra ecosystem in the eastern Canadian High-Arctic.An important biological characteristic of the area is the presence of a large snow goose (scientific names of most vertebrate species can be found in \textbf{Table \ref{table:species_name_strategy}}) colony of around 25 000 breeding pairs \citep{reed2002} spanning approximately 70 km\textsuperscript{2}. The vertebrate community within the study area comprises 30 bird species, with 29 of them being migratory or partially migratory, along with 5 mammal species (\textbf{Table \ref{table:species_name_strategy}}; \citet{moisan2023, gauthier2024a}). The study area experiences significant temporal fluctuations in the population of small mammals (lemmings), which in turn impact the occurrence and abundance of their avian and mammalian predators such as snowy owls, rough-legged hawks, long-tailed jaegers, ermines and Arctic foxes \citep{legagneux2012}. We exclude occasional visitors, namely: i) species lacking confirmed breeding occurrences on the study site, ii) species observed solely within a single year, and iii) species primarily breeding and foraging in nearby marine or coastal habitats \citep{moisan2023}. The case of the red fox (\textit{Vulpes vulpes}) was ambiguous. While the presence of a breeding pair has been confirmed in the study area \citep{lai2022}, the extent of population establishment remains unclear and sightings are rare. Therefore, we decided to exclude this species.

         		\input{tables/species_name_strategy.tex}
%label-> table:species_name_strategy
\newpage
            	\subsubsection*{b. Geography} Our 463 km\textsuperscript{2} study area is located on the southern plain of Bylot Island, Nunavut, Canada (72.889 N, -79.906 W; \textbf{Figure \ref{figure:zones}}).
            	\subsubsection*{c. Habitat}  The study area comprises a combination of mesic tundra mainly on hills (64 \%), upland plateaus of sedimentary rock with drier/rockier habitat at higher elevation (20 \%), low-lying wetlands interspersed with ponds (10 \%) and larger bodies of water such as lakes and rivers (6 \%).
            	\subsubsection*{d. Geology} see \citet{klassen1993} for a complete and detailed description of the geology of the study area.
            	\subsubsection*{e. Hydrology} Wetlands were delineated by photo-interpretation of high-resolution satellite images (30 cm; Ouellet, unpublished data), whereas lakes were delineated with aerial photos and rivers with google satellite images, resulting in a coarser delineation.
            	\subsubsection*{f. Site history} see \citet{gauthier2024a, gauthier2024b}  for a complete and detailed history of the site.
            	\subsubsection*{g. Climate} The mean temperature in July is 6°C, and the study area typically remains free of snow from mid-June to late September \citep{gauthier2013}. The climate of the southern plain of Bylot Island is generally milder than that of the surrounding latitudes, as the plain present a southern exposure and the mountains to the north protect the plain from cold northerly winds. \citep{gauthier2024a}.
 \newpage

      	\subsection*{2. Experimental or sampling design}
        		\subsubsection*{a. Permanent plots}
The study area is divided into 9 zones based on the sampling method and the level of field effort applied in each zone (\textbf{Figure \ref{figure:zones}}). Long-term monitoring of the community began in the 1990s in the Qarlikturvik valley \citep{gauthier2013, gauthier2024a}, which represents the zone of the study area with the highest annual sampling effort. Within the Qarlikturvik valley, the sampling is concentrated on the southern side of the glacial river (\textbf{Figure \ref{figure:qarlikturvik_valley}}), where the main research infrastructure is located. Another zone with extensive sampling efforts is Camp 2, located at the core of the snow goose colony, where the primary focus is to monitor snow goose nests. However, nests of many other avian species are also monitored within and around the colony in this zone. Camp 3, Pointe Dufour, Goose Point, and Malaview are zones where intensive sampling efforts are conducted annually, albeit for a relatively brief period (approximately one week) during the breeding season of most species \citep{gauthier2024a}. The upland zones in the study area (defined as areas approximately 300 meters above sea level or more) are the Black Plateau, Southern Plateau, and Camp 3 Plateau. These zones are primarily visited to assess raptor nesting activity \citep{beardsell2016}. The zone between the Qarlikturvik valley and Camp 3 received very little sampling effort and is therefore excluded from the study area.

\begin{figure}
\centering
  \includegraphics[width=0.8\textwidth, angle=0]{figures/zones_study_area.pdf}
   \caption{Map of the different zones (colored polygons) of the 463 km\textsuperscript{2} study area located on the southern plain of Bylot Island, Nunavut Canada. The outline of the snow goose colony is presented with white dash; we used the outline of the colony in 2015 since it represents an average colony area (67 km\textsuperscript{2}).}
  \label{figure:zones}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.8\textwidth, angle=0]{figures/qarlikturvik_valley.pdf}
  \vspace{-70pt} % Reduce space between figure and caption
  \caption{Intensive nests searching plots (8 km\textsuperscript{2} and 2 km\textsuperscript{2}) located in the Qarlikturvik valley.}
 \label{figure:qarlikturvik_valley}
\end{figure}
\newpage

		\subsubsection*{b. Avian nest monitoring}
Avian nest monitoring was conducted annually except in 2020 and 2021 due to logistical constraints imposed by the COVID-19 pandemic. Systematic nest monitoring refers here to a systematic sampling approach aimed at documenting all nests within a specified area. Monitoring is considered opportunistic when there is a chance that some nests might not have been detected within a specific area. Nest densities derived from nest sampling could be underestimated due to early nest failure (i.e., failure that happened before our sampling period).

\input{tables/species_year_monitoring.tex}
%label-> table:species_year_monitoring

	\begin{enumerate}[label=\roman*]
	\item[] \textit{\textbf{Pacific loon, red-throated loon, cackling goose, tundra swan and glaucous gull}}\\
	Since 2004, systematic searches of wetland areas have been conducted on the southern side of the glacial river in the Qarlikturvik Valley, and since 2017, in other zones of the study area. This sampling aimed to find all nests of the cackling goose and the glaucous gull. Nest locations of other large wetland-nesting species, including the tundra swan, the red-throated loon and the Pacific loon, were also noted, as these species nest in the same habitat \citep{duchesne2021,gauthier2024a}. Each year, all known or potential nesting sites were revisited. Observers detected nests by walking and scanning around ponds and lakeshores to identify any active nesting sites. These large species can be seen from a relatively long distance sitting on the nest or when flushing from the nest. Most of them (geese, swans and gulls) can also reveal their presence with alarm calls or nest defense displays. We are confident that nest detection probability was high for these species given the open landscape.\\

	\item[] \textit{\textbf{Snow goose}}\\
	Snow geese nest in a large colony in the study area (\textbf{Figure \ref{figure:zones}}), but also in small aggregations distributed on the island, especially in years when snowy owls are nesting \citep{lepage1996,reed2002}. Since 1994, goose nests were systematically monitored on a 0.24 km\textsuperscript{2} wetland at the center of the colony. Since 1999, nests were also systematically monitored on a variable number of plots, measuring 0.01 km\textsuperscript{2} in wetland habitat and 0.04 km\textsuperscript{2} in mesic habitat, randomly distributed throughout the goose colony \citep{gauthier2020goose}. The total area covered by the randomly distributed plots averaged 0.79 \textpm{0.37} km\textsuperscript{2} per year. From 2010 onwards, except in 2020 and 2021, we traced the approximate boundary of the goose colony using a GPS receiver aboard a helicopter flying along the colony border \citep{duchesne2021}.

	\item[] \textit{\textbf{Rough-legged hawk, peregrine falcon and common raven}}\\
	Peregrine falcons, rough-legged hawks and common ravens nest on cliffs, near ravines, and on large rocky outcrops and tend to reuse the same nesting sites from one year to the next \citep{beardsell2016}. Systematic monitoring of every known or potential nesting site has been carried out in the Qarlikturvik valley, Black plateau and Southern plateau since 2007 and throughout the study area since 2013 \citep{beardsell2016, gauthier2020avianpred}. Observers walked along ridges and scanned surrounding areas from vantage points to detect nesting birds. These large species can be seen from a relatively long distance sitting on the nest or when flushing from the nest. They can also reveal their presence with alarm calls or nest defense displays. We are confident that nest detection probability was high for these species. Each year the observers use slightly different paths to sample the areas, but locate the nests in the same positions, which supports a high probability of detection for these species. Most nesting sites were located in the upland zones of the study area, which include the Black Plateau, Southern Plateau and Camp 3 Plateau.

	\item[] \textit{\textbf{Snowy owl}} \\
	Snowy owls predominantly nest in habitats similar to other raptors, favoring ridges in mountainous or hilly regions, although they can occasionally be found nesting on mounds in lowland areas \citep{seyer2020}. Since 1996, searches for snowy owl nests have been conducted concurrently with searches for other raptor nests in the Black and Southern plateaus, as well as during searches for jaeger nests on the southern side of the glacial river in the Qarlikturvik Valley. Additionally, since 2012, nests have been recorded across the entire study area by scanning the landscape from hills and ridges during the nesting period \citep{duchesne2021}. Given that snowy owls nest on elevated mounds, exhibit contrasting colors with the landscape, emit alarm calls, and display defensive behaviors, active nesting sites have a high probability of detection.\\
